Long-form recording of infant body position in the home using wearable inertial sensors

### 2023-08-14 manuscript.Rmd
Working on revision

# SUBMITTED DATA IS BASED ON THE FOLLOWING SESSION IDS
#9908  9909  9910  9911 10201 10203 10204 10301 10501 10601 10701 10702 10703 10704 10802 10901 11001 11002 11004 11102 11103 11104 11202 11501 11601 11901 12001 12002 12101 12302 12303 12501 12502 13001
```

@@

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. Data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories. Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is complex because movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in "tummy time", or they can crawl in a myriad of ways [@AdolphVereijken1998]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, or in a "W" position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [@KwonZavos2019; @PatelShi2019].

@@

```{r participants, cache=TRUE, cache.extra=tools::md5sum("data/imu-compiled.csv")}
ppts <- read_csv("data/imu-compiled.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older"))) %>% 
  filter(id_uni != 10401)
num_younger <- ppts %>% filter(group == "Younger") %>% .$id %>% unique(.) %>% length(.)
num_older <-  ppts %>% filter(group == "Older") %>% .$id %>% unique(.) %>% length(.)
num_sessions <- ppts %>% group_by(group, id, session) %>% slice_head(n = 1) %>% count %>% group_by(group, id) %>% count
age <- ppts %>% group_by(group) %>% get_summary_stats(age) %>% column_to_rownames("group")
```

Infants were recruited in one of two age groups: *Younger* infants were between 4 and 7 months and *older* infants were between 11 and 14 months. There were `r num_younger` infants in the 4-7 month group and `r num_older` in the 11-14 month group. Ten infants were female and 12 were male. Families were recruited through social media advertisements and from community events in Southern California. Most infants were reported to be either Hispanic and White or Non-Hispanic and White. Families were compensated \$30 for every home recording session they completed. The BLINDED Institutional Review Board reviewed and approved all procedures associated with the study. All caregivers gave their informed consent before the start of the study.

Most participants were tested in a single session (*n* = `r nrow(num_sessions %>% filter(n == 1))`), but `r nrow(num_sessions %>% filter(n > 1))` participants contributed between 2-4 sessions as part of an ongoing longitudinal study. Although including more data from some participants could over-represent their characteristics in the model, we reasoned that this drawback was outweighed by having more available data to use for training. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of `r length(unique(ppts$id_uni))` sessions, with `r num_sessions %>% filter(group == "Younger") %>% pull(n) %>% sum` sessions from younger infants and `r num_sessions %>% filter(group == "Older") %>% pull(n) %>% sum` sessions from older infants. Across sessions, younger infants' age ranged from `r apa_num(age["Younger","min"], digits = 1)` to `r apa_num(age["Younger","max"], digits = 1)` months (*M* = `r apa_num(age["Younger","mean"], digits = 1)`) and older infants' age ranged from `r apa_num(age["Older","min"], digits = 1)` to `r apa_num(age["Older","max"], digits = 1)` months (*M* = `r apa_num(age["Older","mean"], digits = 1)`). The total number of recording sessions (`r length(unique(ppts$id_uni))`) exceeded the number of sessions employed in comparable past work: 10 in @NamPark2013, 15 in @FranchakScott2021, 22 in @AiraksinenRasanen2020, and 33 in @YaoPlotz2019.

@@

As in @AiraksinenRasanen2020, overall accuracy decreased when fewer sensors were used. Table \@ref(tab:metrics-reduce) compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model *M* = `r metrics_sensor["All", "mean_group"]`; individual model *M* = `r metrics_sensor["All", "mean_split"]`), and the lowest when using only a single sensor (left ankle group model *M* = `r metrics_sensor["Left Ankle", "mean_group"]`; right ankle individual model *M* = `r metrics_sensor["Right Ankle", "mean_split"]`). Accuracy for some pairs approached 4-sensor accuracy: Left ankle and thigh had group model accuracy of *M* = `r metrics_sensor["Left Thigh/Ankle", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Left Thigh/Ankle", "mean_split"]`, only 2.2% and 1.5% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_split"]`, 12.7% and 7.2% worse than using all four sensors.

@@

age_scatter <- function(ds) {
  ds %>% drop_na(prop) %>% 
  ggplot() + facet_wrap("position", as.table = T, ncol = 2) + 
  geom_smooth(aes(x = age, y = prop), color = "#191919", method = "lm", se = T) + 
  stat_summary(fun = "mean", mapping = aes(x = age, y = prop, group = id_uni, fill = position), shape = 21, size = .7) + 
  scale_fill_manual(values = pal, guide = "none") + 
  scale_x_continuous(name = "Age (months)", breaks = seq(3, 15, 3), limits = c(3,15)) + 
  ylim(0,1) + ylab("Proportion of waking day") + 
  theme(panel.spacing.x = unit(.5, "lines")) +
  theme(panel.spacing.y = unit(1, "lines")) + 
  theme(strip.background = element_blank(), strip.text.x = element_text(size = 16, face = "bold"))
}

>>>>

# SUBMITTED DATA IS BASED ON THE FOLLOWING SESSION IDS
#9908  9909  9910  9911 10201 10203 10204 10301 10501 10601 10701 10702 10703 10704 10802 10901 11001 11002 11004 11102 11103 11104 11202 11501 11601 11901 12001 12002 12101 12302 12303 12501 12502 13001
# repeat sessions - younger = 3
# repeat sessions - older = 4
```

@@

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. Data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories. Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is complex because movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in "tummy time", or they can crawl in a myriad of ways [@AdolphVereijken1998]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, or in a "W" position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat likely restricts the range of torso orientations compared with sitting independently or on a caregiver's lap. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [@KwonZavos2019; @PatelShi2019].

@@

```{r participants, cache=TRUE, cache.extra=tools::md5sum("data/imu-compiled.csv")}
ppts <- read_csv("data/imu-compiled.csv") %>% 
  mutate(id_uni = factor(id*100+session),
         group = factor(ifelse(age <= 8, "Younger", "Older"))) %>% 
  filter(id_uni != 10401)
num_younger <- ppts %>% filter(group == "Younger") %>% .$id %>% unique(.) %>% length(.)
num_older <-  ppts %>% filter(group == "Older") %>% .$id %>% unique(.) %>% length(.)
num_sessions <- ppts %>% group_by(group, id, session) %>% slice_head(n = 1) %>% count %>% group_by(group, id) %>% count
age <- ppts %>% group_by(group) %>% get_summary_stats(age) %>% column_to_rownames("group")

Infants were recruited in one of two age groups: *Younger* infants were between 4 and 7 months and *older* infants were between 11 and 14 months. There were `r num_younger` infants in the 4-7 month group and `r num_older` in the 11-14 month group. Ten infants were female and 12 were male. Families were recruited through social media advertisements and from community events in Southern California. Most infants were reported to be either Hispanic and White (*n* = 8) or Non-Hispanic and White (*n* = 7). Families were compensated \$30 for every home recording session they completed. The BLINDED Institutional Review Board reviewed and approved all procedures associated with the study. All caregivers gave their informed consent before the start of the study.

Most participants were tested in a single session (*n* = `r nrow(num_sessions %>% filter(n == 1))`), but `r nrow(num_sessions %>% filter(n > 1))` participants contributed between 2-4 sessions as part of an ongoing longitudinal study (three from the younger group and four from the older group). Although including more data from some participants could over-represent their characteristics in the model, we reasoned that this drawback was outweighed by having more available data to use for training. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of `r length(unique(ppts$id_uni))` sessions, with `r num_sessions %>% filter(group == "Younger") %>% pull(n) %>% sum` sessions from younger infants and `r num_sessions %>% filter(group == "Older") %>% pull(n) %>% sum` sessions from older infants. Across sessions, younger infants' age ranged from `r apa_num(age["Younger","min"], digits = 1)` to `r apa_num(age["Younger","max"], digits = 1)` months (*M* = `r apa_num(age["Younger","mean"], digits = 1)`) and older infants' age ranged from `r apa_num(age["Older","min"], digits = 1)` to `r apa_num(age["Older","max"], digits = 1)` months (*M* = `r apa_num(age["Older","mean"], digits = 1)`). The total number of recording sessions (`r length(unique(ppts$id_uni))`) exceeded the number of sessions employed in comparable past work: 10 in @NamPark2013, 15 in @FranchakScott2021, 22 in @AiraksinenRasanen2020, and 33 in @YaoPlotz2019.

@@

As in @AiraksinenRasanen2020, overall accuracy decreased when fewer sensors were used. Table \@ref(tab:metrics-reduce) compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model *M* = `r metrics_sensor["All", "mean_group"]`; individual model *M* = `r metrics_sensor["All", "mean_split"]`), and the lowest when using only a single sensor (left ankle group model *M* = `r metrics_sensor["Left Ankle", "mean_group"]`; right ankle individual model *M* = `r metrics_sensor["Right Ankle", "mean_split"]`). Accuracy for some pairs approached 4-sensor accuracy: Left ankle and thigh had group model accuracy of *M* = `r metrics_sensor["Left Hip/Ankle", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Left Hip/Ankle", "mean_split"]`, only 2.2% and 1.5% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_group"]` and individual model accuracy of *M* = `r metrics_sensor["Both Ankles", "mean_split"]`, 12.7% and 7.2% worse than using all four sensors.

@@

age_scatter <- function(ds) {
  ds %>% drop_na(prop) %>% 
  ggplot() + facet_wrap("position", as.table = T, ncol = 2, scales = "free_x") + 
  geom_smooth(aes(x = age, y = prop), color = "#191919", method = "lm", se = T) + 
  stat_summary(fun = "mean", mapping = aes(x = age, y = prop, group = id_uni, fill = position), shape = 21, size = .7) + 
  scale_fill_manual(values = pal, guide = "none") + 
  scale_x_continuous(name = "Age (months)", breaks = seq(3, 15, 3), limits = c(3,15)) + 
    scale_y_continuous(name = "Proportion of waking day", breaks = seq(0, 1, .25), limits = c(0,1)) + 
  theme(panel.spacing.x = unit(.5, "lines")) +
  theme(panel.spacing.y = unit(1, "lines")) + 
  theme(strip.background = element_blank(), strip.text.x = element_text(size = 16, face = "bold"))
}

### 2023-08-14 manuscript.tex
Working on revision

The open question is whether \emph{automatic classification} is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. Data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories. Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is complex because movement can vary greatly \emph{within} a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in ``tummy time'', or they can crawl in a myriad of ways \autocite{AdolphVereijken1998}. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a ``V'' position with legs fully extended, or in a ``W'' position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity \autocite{KwonZavos2019,PatelShi2019}.

@@

Infants were recruited in one of two age groups: \emph{Younger} infants were between 4 and 7 months and \emph{older} infants were between 11 and 14 months. There were 8 infants in the 4-7 month group and 14 in the 11-14 month group. Ten infants were female and 12 were male. Families were recruited through social media advertisements and from community events in Southern California. Most infants were reported to be either Hispanic and White or Non-Hispanic and White. Families were compensated \$30 for every home recording session they completed. The BLINDED Institutional Review Board reviewed and approved all procedures associated with the study. All caregivers gave their informed consent before the start of the study.

Most participants were tested in a single session (\emph{n} = 15), but 7 participants contributed between 2-4 sessions as part of an ongoing longitudinal study. Although including more data from some participants could over-represent their characteristics in the model, we reasoned that this drawback was outweighed by having more available data to use for training. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of 34 sessions, with 14 sessions from younger infants and 20 sessions from older infants. Across sessions, younger infants' age ranged from 3.8 to 7.2 months (\emph{M} = 5.2) and older infants' age ranged from 10.7 to 14.2 months (\emph{M} = 11.7). The total number of recording sessions (34) exceeded the number of sessions employed in comparable past work: 10 in \textcite{NamPark2013}, 15 in \textcite{FranchakScott2021}, 22 in \textcite{AiraksinenRasanen2020}, and 33 in \textcite{YaoPlotz2019}.

@@

As in \textcite{AiraksinenRasanen2020}, overall accuracy decreased when fewer sensors were used. Table \ref{tab:metrics-reduce} compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model \emph{M} = 0.846; individual model \emph{M} = 0.916), and the lowest when using only a single sensor (left ankle group model \emph{M} = 0.667; right ankle individual model \emph{M} = 0.819). Accuracy for some pairs approached 4-sensor accuracy: Left ankle and thigh had group model accuracy of \emph{M} = NA and individual model accuracy of \emph{M} = NA, only 2.2\% and 1.5\% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of \emph{M} = 0.719 and individual model accuracy of \emph{M} = 0.844, 12.7\% and 7.2\% worse than using all four sensors.

>>>>

The open question is whether \emph{automatic classification} is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. Data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories. Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is complex because movement can vary greatly \emph{within} a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in ``tummy time'', or they can crawl in a myriad of ways \autocite{AdolphVereijken1998}. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a ``V'' position with legs fully extended, or in a ``W'' position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat likely restricts the range of torso orientations compared with sitting independently or on a caregiver's lap. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity \autocite{KwonZavos2019,PatelShi2019}.

@@

Infants were recruited in one of two age groups: \emph{Younger} infants were between 4 and 7 months and \emph{older} infants were between 11 and 14 months. There were 8 infants in the 4-7 month group and 14 in the 11-14 month group. Ten infants were female and 12 were male. Families were recruited through social media advertisements and from community events in Southern California. Most infants were reported to be either Hispanic and White (\emph{n} = 8) or Non-Hispanic and White (\emph{n} = 7). Families were compensated \$30 for every home recording session they completed. The BLINDED Institutional Review Board reviewed and approved all procedures associated with the study. All caregivers gave their informed consent before the start of the study.

Most participants were tested in a single session (\emph{n} = 15), but 7 participants contributed between 2-4 sessions as part of an ongoing longitudinal study (three from the younger group and four from the older group). Although including more data from some participants could over-represent their characteristics in the model, we reasoned that this drawback was outweighed by having more available data to use for training. Only 1 session was excluded due to a technical error---one of the four IMU sensors failed to record, resulting in an unusable set of data for classification. Across the two age groups, we report data on a total of 34 sessions, with 14 sessions from younger infants and 20 sessions from older infants. Across sessions, younger infants' age ranged from 3.8 to 7.2 months (\emph{M} = 5.2) and older infants' age ranged from 10.7 to 14.2 months (\emph{M} = 11.7). The total number of recording sessions (34) exceeded the number of sessions employed in comparable past work: 10 in \textcite{NamPark2013}, 15 in \textcite{FranchakScott2021}, 22 in \textcite{AiraksinenRasanen2020}, and 33 in \textcite{YaoPlotz2019}.

@@

As in \textcite{AiraksinenRasanen2020}, overall accuracy decreased when fewer sensors were used. Table \ref{tab:metrics-reduce} compared overall accuracy for all four sensors (top row), pairs of 2 sensors (rows 2-5), and single sensors (rows 6-9). The highest accuracy was observed when using all four sensors (group model \emph{M} = 0.846; individual model \emph{M} = 0.916), and the lowest when using only a single sensor (left ankle group model \emph{M} = 0.667; right ankle individual model \emph{M} = 0.819). Accuracy for some pairs approached 4-sensor accuracy: Left ankle and thigh had group model accuracy of \emph{M} = 0.824 and individual model accuracy of \emph{M} = 0.901, only 2.2\% and 1.5\% worse than using all four sensors. Other pairings were less accurate; notably, using both ankles resulted in group model accuracy of \emph{M} = 0.719 and individual model accuracy of \emph{M} = 0.844, 12.7\% and 7.2\% worse than using all four sensors.
